---
title: 'Prediction Assignment'
author: "Xiaoyang Jin"
date: "`r Sys.Date()`"
output:
    html_document:
        toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

Using devices such as _Jawbone Up_, _Nike FuelBand_, and _Fitbit_ it is now
possible to collect a large amount of data about personal activity
relatively inexpensively. These type of devices are part of the quantified
self movement – a group of enthusiasts who take measurements about
themselves regularly to improve their health, to find patterns in their
behavior, or because they are tech geeks. One thing that people regularly do
is quantify how _much_ of a particular activity they do, but they rarely
quantify _how well they do it_. In this project, your goal will be to use
data from accelerometers on the belt, forearm, arm, and dumbell of
6 participants. They were asked to perform barbell lifts correctly and
incorrectly in 5 different ways. More information is available from the
website here: <http://groupware.les.inf.puc-rio.br/har> (see the section on
the Weight Lifting Exercise Dataset).

The training data for this project are available here:

<https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv>

The test data are available here:

<https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv>

## Train and select models

### preprocess

First we will read in the data, the training and testing data are already downloaded into the 'data/' directory.
```{r}
training <- read.csv("data/pml-training.csv",
                     na.strings = c("NA", ""),
                     stringsAsFactors = FALSE)
testing <- read.csv("data/pml-testing.csv",
                     na.strings = c("NA", ""),
                     stringsAsFactors = FALSE)
```

```{r}
dim(training)
table(training$classe)
```

As we see, the training dataset has a sample size of `r nrow(training)`,
and there are `r ncol(training)-1` features in this dataset (that's somehow a
huge number compared to the sample size). The **classe** variable is what we 
would predict in the test set, and it seems balanced.

The data type of each feature will be estimated, lest it's misjudged in
`read.csv` function. (In fact, this step shows no effect when I inspect the
change happened)
```{r}
for(i in seq(ncol(training)-1)) {
    vector_valid = training[,i][!is.na(training[,i]) & training[,i] != ""]
    if(length(vector_valid) == 0 ||
       sum(is.na(suppressWarnings(as.numeric(vector_valid)))) > 0) {
        training[,i] = as.character(training[,i])
        testing[,i] = as.character(testing[,i])
    } else {
        training[,i] = as.numeric(training[,i])
        testing[,i] = as.numeric(testing[,i])
    }
}
```

The first 7 columns will be removed because these variables shouldn't relate 
to the result variable. Besides, as the number of features is large, we could safely remove all the features with missing values.
```{r}
training = training[,-c(1:7)]
testing = testing[,-c(1:7)]

column.complete = complete.cases(t(training))
training = training[, column.complete]
testing = testing[, column.complete]

dim(training)
```

After this process, `r dim(training)[2]-1` features remain in our dataset
(the `training` data frame has `r dim(training)[2]` columns because one of
them is the result variable).

Next the training set is cut apart into the de facto training part and the
validation part. The validation part will be used to predict out of sample 
error before we turn to the real test set.
```{r}
library(caret)
set.seed(2017)

train.idx = createDataPartition(training$classe, p = 3/4, list = FALSE)
train_dat = training[train.idx,]
validate_dat = training[-train.idx,]
```

### train model using rpart

Now we will train a decision tree model using `rpart` mothod. 5-fold cross
validation is used in `train` function.
```{r}
library(rpart)
set.seed(2017)

model1 <- train(classe ~ ., data = train_dat,
                method = "rpart",
                trControl = trainControl(method = "cv", number = 5))
```

The final tree model could be plotted using the `fancyRpartPlot` function in
`rattle` package:

```{r}
library(rattle)
fancyRpartPlot(model1$finalModel)
```

The out of sample error could be calculated on the validation set
```{r}
(model1_conf = 
     confusionMatrix(predict(model1, newdata = validate_dat),
                     validate_dat$classe))
```

The overall accuracy is `r model1_conf$overall[1]`

### train model using random forest

Alternatively, we could use the random forset model, which is the default
method in the `train` function

```{r}
library(randomForest)

set.seed(2017)
model2 <- train(classe ~ ., data = train_dat,
                trControl = trainControl(method = "cv", number = 5))
```

Again, we assess the model performance on the validation set
```{r}
(model2_conf =
    confusionMatrix(predict(model2, validate_dat),
                    validate_dat$classe))
```

The accuracy is `r model2_conf$overall[1]`, which is much higher than the
`rpart` method, so we will use the random forest method on the test set.

We could inspect the variable importance of model2 (only the 10 most 
important variables are listed):
```{r}
library(dplyr)
library(ggplot2)
library(ggthemes)

# Get importance
importance    <- importance(model2$finalModel)
varImportance <- data.frame(Variables = row.names(importance), 
                            Importance =
                                round(importance[,'MeanDecreaseGini'],2))

# Create a rank variable based on importance
rankImportance <- varImportance %>%
    top_n(10, Importance) %>%
    mutate(Rank = paste0('#',dense_rank(desc(Importance))))

# Use ggplot2 to visualize the relative importance of variables
ggplot(rankImportance, aes(x = reorder(Variables, Importance), 
    y = Importance, fill = Importance)) +
  geom_bar(stat='identity') + 
  geom_text(aes(x = Variables, y = 0.5, label = Rank),
    hjust=0, vjust=0.55, size = 4, colour = 'red') +
  labs(x = 'Variables') +
  coord_flip() + 
  theme_few()
```

The first two most important variables are `roll_belt` and `pitch_forearm`,
it's in accordance with the previous tree model.

## Predict on testing set
Now we could apply the model on the testing set
```{r}
predict(model2, newdata = testing)
```

## Session information
```{r}
devtools::session_info()
```

